The persuasion of symbolic elements in government public communications, and their moderation by personal relevance

### 2018-07-15 paper3_relevance and symbols_01.Rmd
Add files via upload

Persuasion through symbolic communications and branding can be perceived favorably, insofar as it enables governments to engage more effectively with relevant publics, thereby to increase their trust and enhance their collaboration. The effectiveness of symbols, as opposed to communication of "objective" information, stems from their potency to overcome citizens' negative biases. On the other hand, employing a broader perspective of democratic theory, one can argue that in the public-sector domain, the emotive effect of symbols undermines the fundamental normative expectation that citizens' judgments of government should be based on their critical judgment. Furthermore, as previous studies demonstrated, positive symbolic messages can compensate for poor performance and create a misalignment between citizens' positive perceptions and the "real world" [@alonbarkat_2017; @alonbarkat_2018]. Accordingly, symbols can mitigate justified public criticism of government failures.

In this study I explore the boundaries of governments' persuasion of citizens through symbolic communication. Namely, I seek to explore the conditions under which symbolic elements are likely to be more, or less, effective in shaping (and possibly manipulating) citizens' trust. Building on the social psychology Elaboration Likelihood Model (ELM) [@petty_1986; @petty_2011], I hypothesize that citizens are less susceptible to symbols the more they perceive the policy issue in the communication as having personal relevance for themselves, which enhances their motivation to engage in elaborate processing of information. Similarly, high perceived personal relevance is expected to enhance citizens' inclination to scrutinize the substantive content of communication.

Employing a survey experiment, I test these hypotheses in the context of the Israeli Environment Protection Ministry, and its policy plan for reducing the air pollution in a specific area (the Haifa-Bay). I operationalize the perceived personal relevance of the policy plan both experimentally, and by comparing between citizens from the polluted Haifa-Bay area and others. contrary to my expectations, the results indicate that even high levels of perceived personal relevance do not attenuate the effect of symbols on citizens' trust. These findings may suggest that the boundaries of the effect of symbolic communication are wider than expected. I discuss the theoretical and normative implications of these findings.

@@

The Elaboration Likelihood Model (ELM), developed in the late 1970s by social psychology scholars Richard Petty and John Cacioppo, provides a theoretical framework for investigating the effects of "persuasive" communications on people's attitudes [@petty_1986; @petty_2011]. According to ELM, people may respond differently to a communication, depending on the amount of thinking that they dedicate to it. When people have high motivation and ability to elaborate on a message, they are likely to focus on the arguments in the communication that are central to the merits of the issue, and scrutinize them ("central route"). Conversely, when they are unmotivated or unable to elaborate, they may rely on available heuristic cues that require relatively little amount of cognitive effort ("peripheral route"). These two routes for persuasion are regarded as two ends of an "elaboration continuum". People may also be located between these two edges, and then a mixture of these processes will determine their persuasion.

Viewed against the above theoretical framework of ELM, substantive information in government public communications represent message arguments, which can be critically evaluated based on their logical quality (central route). Conversely, symbolic elements can be regarded as potential peripheral cues, which can affect attitudes by evoking emotions and associations attributed to government organizations and their policies (peripheral route). Accordingly, the more citizens are motivated to think about the content of the communication, and the more capable they are of doing so, the more they are likely to form their attitudes based on substantive information. Conversely, the less motivated and capable they are to elaborate, the more susceptible they are to symbolic elements in the communication.

@@

In this paper, the above-mentioned hypotheses are tested via a randomized survey experiment, building on the methodology of @petty_1983. Focusing on the case of the Israeli Environment Protection Ministry (hereafter: EPM), I examine citizens' responses to communications of EPM regarding two of its current policy plans. The first policy regards *reducing the the air pollution in the highly polluted Haifa-Bay area*, located in the north of Israel. This policy issue offers a fortunate case for examining the moderating role of perceived personal relevance. The distinction between those who reside in the polluted area and those who do not represents natural variation in personal relevance. Moreover, because this policy issue is intangible, ambiguous and complicated, the differences in personal relevance are not strongly confounded with differences in personal knowledge, as is the case in many other policy issues [@petty_1983]. Congruently, those who reside in the polluted Haifa-Bay area are more motivated to scrutinize a communication about the air-pollution in their area of residence, yet no more capable to do so than those who reside elsewhere. As explained below, the sample population of the study consists of residents of the Haifa-bay area, matched with residents of cities at the center of Israel with similar aggregate socio-economic profile. The second policy plan regards *reducing domestic waste and increasing recycling*. This topic is assumed to be equally relevant for Haifa-bay residents and others. Therefore, adding this policy plan is intended to confirm that respondents’ different responses to the air-pollution policy, if found, may be attributed to perceived personal relevance. I henceforth refer to these two policy plans as "air-pollution" and "recycling".

@@

This study contributes to the growing public administration research on citizens' responses to government public communications, and specifically to symbolic elements entangled in these communications. I explored the postulation, derived from social psychology, that citizens are less susceptible to symbols and more likely to scrutinize the substantive information in the communication the more they perceive it as personally relevant. I put this theory to rigorous empirical testing, using a randomized survey experiment focusing on the Israeli Environmental Protection Ministry. I exploited the natural variation in citizens' perceptions of the personal relevance of the ministry’s policy plan regarding air pollution in a specific area.  In addition to utilizing these natural differences, I also manipulated the personal relevance of the policy plans. To the best of my knowledge, this is the first study to assess the moderating role of perceived personal relevance on citizens' responses to government symbolic communications.

Whilst the effect of perceived personal relevance on message processing has been substantiated by experiment-based studies in social psychology and marketing, I did not find support for this hypothesis in the context of citizens' responses to government communications in this study. Subjects residing in the polluted Haifa-Bay indeed reported that they perceive the policy plan for reducing air-pollution in that area as highly relevant for their lives, compared to those who reside in the center. However, despite these differences, Haifa-bay subjects did not pay more attention to the logical quality of the policy plan, and they were equally affected by the symbols in the communication. If anything, the results suggest that the symbols had a greater effect on Haifa-bay subjects. Furthermore, priming subjects to think about the personal relevance of environmental policy, and of the consequences of air-pollution in particular, also failed to moderate the effects of the symbols and information in the communications. These findings are also consistent with the findings of @alonbarkat_2017, where they found that citizens who recently personally experienced prolonged blackouts (and presumably perceived electricity services as personally relevant) were nonetheless significantly affected by a symbol of the government-owned electricity company.

How can we explain the finding that the positive effect of the real symbols was stronger among Haifa-Bay residents? Given that this pattern is also evident for the recycling policy, it is unlikely that it is linked to differences in perceived personal relevance. Rather, they can be explained by the lower baseline of Haifa-Bay subjects' trust in both policies, presumably due to their negative prior attitudes towards EPM. Accordingly, it could be the case that symbolic elements are more effective in shifting people from low to medium levels of trust (i.e. as in the case of Haifa-Bay subjects), than shifting them from medium to high levels (i.e. as in the case of center subjects), indicating a ceiling effect. Stated otherwise, symbols may be more effective in reducing citizens' distrust, as opposed to increasing their trust.[^footnote_ceiling_effect] A related explanation is that the symbols distracted the subjects and decreased their inclination to think about the topic. As a result, they caused Haifa-bay subjects to disregard their prior beliefs about EPM and its failure regarding air-pollution.[^footnote_mediation_elaboration]:

@@

The results of this study, if generalizable, have significant normative implications. They challenge the optimistic view of government direct communications with citizens as a means to inform citizens about government organizations actions and promote transparency and accountability. Instead, they suggest that more often than not, citizens rely on symbols and other peripheral cues in public communications, even when these communications concern issues that have direct implications on their own lives. This inclination is in tension with the assumption that citizens should form their opinions about their government and about public affairs, based on their rational thinking [@dahl_1973; @soroka_2010]. It also buttresses the concern that government organizations will strategically use symbolic communications to manipulate public opinion, and cause citizens to view them more favorably regardless (or even in contrast) to their actual capacities and achievements.

>>>>

Persuasion through symbolic communications and branding can be perceived favorably, insofar as it enables governments to engage more effectively with relevant publics, thereby to increase their trust and enhance their collaboration. The effectiveness of symbols, as opposed to communication of "objective" information, stems from their potency to overcome citizens' negative biases. On the other hand, employing a broader perspective of democratic theory, one can argue that in the public-sector domain, the emotive effect of symbols undermines the fundamental normative expectation that citizens' evaluation of their government should be based on their critical judgment. Furthermore, previous studies have demonstrated that the positive effect of symbols can compensate for organizations' poor performance (@alonbarkat_2017) and for logically unpersuasive informative content (@alonbarkat_2018). Accordingly, symbolic communication can create a misalignment between citizens' positive perceptions and the "real world", and mitigate justified public criticism of government failures.

In this study, I explore the boundaries of governments' persuasion of citizens through symbolic communication, to assess the above concern. Namely, I seek to explore the conditions under which symbolic elements are likely to be more, or less, effective in shaping (and possibly manipulating) citizens' trust. Building on the social psychology Elaboration Likelihood Model (ELM) [@petty_1986; @petty_2011], I theorize that citizens are less susceptible to symbols the more they perceive the policy issue in the communication as having personal relevance for themselves, which enhances their motivation to engage in elaborate processing of information. Similarly, high perceived personal relevance is expected to enhance citizens' inclination to scrutinize the substantive content of communication. Employing a survey experiment, I test these hypotheses in the context of the Israeli Environment Protection Ministry, and its policy plan for reducing the air pollution in a specific area (the Haifa-Bay). I operationalize the perceived personal relevance of the policy plan both experimentally, and by comparing between citizens from the polluted Haifa-Bay area and others. The results indicate that even high levels of perceived personal relevance do not attenuate the effect of symbols on citizens' trust, contrary to my expectations. These findings may suggest that the boundaries of the effect of symbolic communication are wider than expected. I discuss the theoretical and normative implications of these findings.

@@

The Elaboration Likelihood Model (ELM), developed in the late 1970s by social psychology scholars Richard Petty and John Cacioppo, provides a theoretical framework for investigating the effects of communications on people's attitudes [@petty_1986; @petty_2011]. According to ELM, people may respond differently to a communication, depending on the amount of thinking that they dedicate to it. When people have high motivation and ability to elaborate on a message, they are likely to focus on the arguments in the communication that are central to the merits of the issue, and scrutinize them ("central route"). Conversely, when they are unmotivated or unable to elaborate, they may rely on available heuristic cues that require relatively little amount of cognitive effort ("peripheral route"). These two routes for persuasion are regarded as two ends of an "elaboration continuum". People may also be located between these two edges, and then a mixture of these processes will determine their persuasion.

Viewed against the above theoretical framework, substantive information in government public communications represent message arguments, which can be critically evaluated based on their logical quality (central route). Conversely, symbolic elements can be regarded as potential peripheral cues, which can affect attitudes by evoking emotions and associations attributed to government organizations and their policies (peripheral route). Accordingly, the more citizens are motivated to think about the content of the communication, and the more capable they are of doing so, the more they are likely to form their attitudes based on substantive information. Conversely, the less motivated and capable they are to elaborate, the more susceptible they are to symbolic elements in the communication.

@@

In this paper, the above-mentioned theoretical expectations are tested via a randomized survey experiment, building on the methodology of @petty_1983. Focusing on the case of the Israeli Environment Protection Ministry (hereafter: EPM), I examine citizens' responses to communications of EPM regarding two of its current policy plans. The first policy regards *reducing the the air pollution in the highly polluted Haifa-Bay area*, located in the north of Israel. This policy issue offers a fortunate case for examining the moderating role of perceived personal relevance. The distinction between those who reside in the polluted area and those who do not represents natural variation in personal relevance. Moreover, because this policy issue is intangible, ambiguous and complicated, the differences in personal relevance are not strongly confounded with differences in personal knowledge, as is the case in many other policy issues [@petty_1983]. Congruently, those who reside in the polluted Haifa-Bay area are more motivated to scrutinize a communication about the air-pollution in their area of residence, yet no more capable to do so than those who reside elsewhere. As explained below, the sample population of the study consists of residents of the Haifa-bay area, matched with residents of cities at the center of Israel with similar aggregate socio-economic profile. The second policy plan regards *reducing domestic waste and increasing recycling*. This topic is assumed to be equally relevant for Haifa-bay residents and others. Therefore, adding this policy plan is intended to confirm that respondents’ different responses to the air-pollution policy, if found, may be attributed to perceived personal relevance. I henceforth refer to these two policy plans as "air-pollution" and "recycling".

@@

Public communications has been proposed as a remedy for citizens’ low trust in government, and their underestimation of public-service performance. Congruently, governments' investments in public communications, and specifically in branding and advertising, has increased over the past generation in many economically-developed democratic countries. This study contributes to the growing public administration research on citizens' responses to government public communications, and specifically to symbolic elements entangled in these communications. I explored the expectation, derived from social psychology, that citizens are less affected by symbols and more likely to scrutinize the content of the communication the more they perceive it as relevant to their own lives. I put these hypotheses to rigorous empirical testing, using a randomized survey experiment focusing on the Israeli Environmental Protection Ministry. I exploited the natural variation in citizens' perceptions of the personal relevance of the ministry’s policy plan regarding air pollution in a specific area. In addition to utilizing these natural differences, I also manipulated the personal relevance of the policy plans. To the best of my knowledge, this is the first study to assess the moderating role of perceived personal relevance on citizens' responses to government symbolic communications or branding.

Contrary to my expectations, and those of extant research in social psychology and marketing, I did not find that high levels of perceived personal relevance attenuate the effect of symbols on citizens' trust. Subjects residing in the polluted Haifa-Bay indeed reported that they perceive the policy plan for reducing air-pollution in that area as highly relevant for their lives, compared to those who reside in the center. However, despite these significant differences, Haifa-bay subjects did not pay more attention to the logical quality of the policy plan, and they were equally affected by the symbols in the communication. If anything, the results suggest that the symbols had a greater effect on Haifa-bay subjects. Furthermore, priming subjects to think about the personal relevance of environmental policy, and of the consequences of air-pollution in particular, also failed to moderate the effects of the symbols and information in the communications. These findings are also consistent with the findings of @alonbarkat_2017, where they found that citizens who recently personally experienced prolonged blackouts (and presumably perceived electricity services as personally relevant) were nonetheless significantly affected by a symbol of the government-owned electricity company.

The results of this study, if generalizable, have significant implications. They challenge the optimistic view of government direct communications with citizens as a means to inform citizens about government organizations actions and promote transparency and accountability. Instead, they highlight the significant role played by superficial symbols in public communications in shaping citizens view of governments, as suggested by recent scholars of government branding. I demonstrate that citizens can be persuaded by government branding and symbolic communication even with regard to policy issues that have direct implications on their own lives. These conclusions buttress the concern that government organizations will strategically use symbolic communications to manipulate public opinion, and cause citizens to view them more favorably regardless (or even in contrast) to their actual capacities and achievements. 
How can we explain the finding that the positive effect of the real symbols was stronger among Haifa-Bay residents? Given that this pattern is also evident for the recycling policy, it is unlikely that it is linked to differences in perceived personal relevance. Rather, they can be explained by the lower baseline of Haifa-Bay subjects' trust in both policies, presumably due to their negative prior attitudes towards EPM. Accordingly, it could be the case that symbolic elements are more effective in shifting people from low to medium levels of trust (i.e. as in the case of Haifa-Bay subjects), than shifting them from medium to high levels (i.e. as in the case of center subjects), indicating a ceiling effect. Stated otherwise, symbols may be more effective in reducing citizens' distrust, as opposed to increasing their trust.[^footnote_ceiling_effect] A related explanation is that the symbols distracted the subjects and decreased their inclination to think about the topic. As a result, they caused Haifa-bay subjects to disregard their prior beliefs about EPM and its failure regarding air-pollution.[^footnote_mediation_elaboration]:

@@
